---
title: "Strategies for analyzing bisulfite-seq data: Comparison MethylSeekR and methylKit on Human ESC methylome"
author: "Alexander Gosdschan"
output: rmarkdown::github_document
date: "`r Sys.Date()`"
---

# Introduction

The analysis of methylation dynamics lead to many descriptions, e.g. a bimodal model of either hyper- or hypomethylated regions dependent on the local density of CpGs (Lövkvist et al., Nucleic Acids Res. 2016). However, given the detection of CpG-poor regions with locally reduced levels of methylation ( on average 30 % ) in pluripotent embryonic stem cells and in neuronal progenitors in both mouse and human, a different model seems also reasonable (Stadler et al., Nature 2011). 

The identification of these low-methylated regions (LMRs) can be achieved by segmentation of the methylome using computational approaches. One of the well-known segmentation methods is based on a three-state Hidden Markov Model (HMM) taking only DNA methylation into account, without knowledge of any additional genomic information such as CpG density or functional annotations (Stadler et al., Nature 2011). Those three states are fully methylated regions (FMRs), unmethylated regions (UMRs) and low-methylated regions (LMRs). This segmentation represents a summary of methylome properties and features, in which unmethylated CpG islands correspond to UMRs (Deaton and Bird, Genes Dev. 2011), the majority is classified as FMR since most of the genome is methylated (Bird A, Genes Dev. 2002) and LMRs represent a new feature with intermediate levels of methylation, poor CpG content and shorter length compared to CpG islands (Stadler et al., Nature 2011). 


# Goal

We compared the results of change-point based segmentation implemented in methylKit to the method used in MethylSeekR based on HMMs but mainly using cutoffs for methylation values. We analysed chromosome 2 of the human H1 embryonic stem cell methylome from the Roadmap Epigenomics Project (Roadmap Epigenomics Consortium et al., Nature 2015). 

# Functions 

Load libraries and functions.

```{r}
library(MethylSeekR)
library("BSgenome.Hsapiens.UCSC.hg19")
library(rtracklayer)
library(genomation)
library(methylKit)
library(reshape2)
library(gridExtra)
library(ggplot2)
library(grid)

source("functions/functions.R")
```

Set global variables for input and output files.

```{r}

data_dir <- "/data/akalin/agosdsc/projects/methylation_paper/data/ESC/"
figure_dir <- "/data/akalin/agosdsc/projects/methylation_paper/figures/"


meth_bw <- "/data/akalin/Base/RoadmapEpigenomics/Experiment/DNAme_WGBS/FractionalMethylation_bigwig/E003_WGBS_FractionalMethylation.bigwig"
cov_bw <- "/data/akalin/Base/RoadmapEpigenomics/Experiment/DNAme_WGBS/ReadCoverage_bigwig/E003_WGBS_ReadCoverage.bigwig"

methtab_file <- paste0(data_dir,"Human_esc_hg19_chr2.tab")
meth.preload_chr2 <- paste0(data_dir,"ESC_WGBS_Methylation_chr2.rds")
meth.preload_full <- paste0(data_dir,"ESC_WGBS_Methylation.rds")
  
chromhmm_file <- paste0(data_dir,"E003_15_coreMarks_mnemonics.bed.gz")
chromhmm_states <- paste0(data_dir,"chromHMM.15states.txt")

```


# Load input data

Download the WGBS data from Roadmap.  

```{r,eval=FALSE}

if(!dir.exists(data_dir) ) dir.create(data_dir,recursive = TRUE)
if(!dir.exists(figure_dir) ) dir.create(figure_dir,recursive = TRUE)

## WGBS methylation data Human H1 ESC

download.meth = function(output.file){
  url="http://egg2.wustl.edu/roadmap/data/byDataType/dnamethylation/WGBS/FractionalMethylation_bigwig/E003_WGBS_FractionalMethylation.bigwig"
  download.file(url, 
              destfile=output.file,
              method="wget",
              quiet = FALSE, mode = "w",
              cacheOK = TRUE)
}

#download.meth(meth_bw)


## WGBS coverage data Human H1 ESC

download.cov = function(output.file){
  url="http://egg2.wustl.edu/roadmap/data/byDataType/dnamethylation/WGBS/ReadCoverage_bigwig/E003_WGBS_ReadCoverage.bigwig"
  download.file(url, 
              destfile=output.file,
              method="wget",
              quiet = FALSE, mode = "w",
              cacheOK = TRUE)
}

#download.cov(cov_bw)

## ChromHMM data Human H1 ESC

download.chromhmm = function(output.file){
  url="http://egg2.wustl.edu/roadmap/data/byFileType/chromhmmSegmentations/ChmmModels/core_K27ac/jointModel/final/all.mnemonics.bedFiles.tgz"
  download.file(url, 
              destfile=paste0(data_dir,"all.mnemonics.bedFiles.tgz"),
              method="wget",
              quiet = FALSE, mode = "w",
              cacheOK = TRUE)
  untar(paste0(data_dir,"all.mnemonics.bedFiles.tgz"),
        files = basename(output.file),exdir = dirname(output.file),
        compressed = "gzip")
  #unlink(paste0(data_dir,"all.mnemonics.bedFiles.tgz"))
  
}

#download.chromhmm(chromhmm_file)

write.chromhmm <- function(output.file) { 
  ## get the true state names from http://egg2.wustl.edu/roadmap/web_portal/chr_state_learning.html#core_15state and store into simple txt file
  
  chromHMM.15states <- 
  "STATE NO.	MNEMONIC	DESCRIPTION	COLOR NAME	COLOR CODE
  1	TssA	Active TSS	Red	255,0,0
  2	TssAFlnk	Flanking Active TSS	Orange Red	255,69,0
  3	TxFlnk	Transcr. at gene 5' and 3'	LimeGreen	50,205,50
  4	Tx	Strong transcription	Green	0,128,0
  5	TxWk	Weak transcription	DarkGreen	0,100,0
  6	EnhG	Genic enhancers	GreenYellow	194,225,5
  7	Enh	Enhancers	Yellow	255,255,0
  8	ZNF/Rpts	ZNF genes & repeats	Medium Aquamarine	102,205,170
  9	Het	Heterochromatin	PaleTurquoise	138,145,208
  10	TssBiv	Bivalent/Poised TSS	IndianRed	205,92,92
  11	BivFlnk	Flanking Bivalent TSS/Enh	DarkSalmon	233,150,122
  12	EnhBiv	Bivalent Enhancer	DarkKhaki	189,183,107
  13	ReprPC	Repressed PolyComb	Silver	128,128,128
  14	ReprPCWk	Weak Repressed PolyComb	Gainsboro	192,192,192
  15	Quies	Quiescent/Low	White	255,255,255"
  
  write(chromHMM.15states,output.file  )
}

#write.chromhmm(output.file = chromhmm_states)

preload.meth <- function(meth.bw, chr=NULL,output.file) {

  meth <- BigWigFile(meth.bw)
  
  chrlens <- seqinfo(meth) # get all chrs
  
  if(is.null(chr)) { len <- seqlengths(chrlens) }
  else { len <- seqlengths(chrlens[chr]) } # length of chr 
  
  mbw <- import(meth, which=GRanges(seqnames=names(len),ranges=IRanges(1,len)))

  saveRDS(mbw,
          file = output.file)
  
}

#preload.meth(meth.bw = meth_bw,chr = "chr2",output.file = meth.preload_chr2)
#preload.meth(meth.bw = meth_bw,chr = NULL,output.file = meth.preload_full)


```


# Preprocess Data

For MethylSeekR we need to create a custom table with those columns: `chromosome position Total-counts Methylation-counts`.
We have a custom function to extract this information from methylation and coverage bigwig `readbw()`. 

```{r, eval=FALSE}

## we create the methylation table for the whole genome
tab <- readbw(methbw = meth_bw,covbw = cov_bw,chr = "chr2")

# for methylSeekR we need chromosome position Total counts Methylation counts
# select columns: chr start coverage numCs
methtab.dt <- tab[,c(1,2,5,6)]
# write output back into text file
write.table(methtab.dt, file = methtab_file,
            quote = FALSE, col.names = FALSE, row.names = FALSE, sep = "\t")

```


## Segment using MethylSeekR

Use MethySeekR functions to get the required segmentation.


```{r,eval=FALSE}
# Run MethylSeekR ---------------------------------------------------------
### here we use the steps described in the vignette to obtain our Segments


# library(MethylSeekR)
set.seed(123)


# biocLite("BSgenome.Hsapiens.UCSC.hg19")
# library("BSgenome.Hsapiens.UCSC.hg19")
sLengths=seqlengths(Hsapiens)
head(sLengths)

## load methylome
methFname <- methtab_file
meth.gr <- readMethylome(FileName=methFname, seqLengths=sLengths)

## detect the PMDs
PMDsegments.gr <- segmentPMDs(m=meth.gr, chr.sel="chr2",
                              seqLengths=sLengths, num.cores=4)

## save PMD ranges
savePMDSegments(PMDs=PMDsegments.gr,
                GRangesFilename=paste0(data_dir,"PMDs.gr.rds"), 
                TableFilename=paste0(data_dir,"PMDs.tab"))



## load CpGislands
# library(rtracklayer)
session <- browserSession()
genome(session) <- "hg19"
query <- ucscTableQuery(session, "cpgIslandExt")
CpGislands.gr <- track(query)
genome(CpGislands.gr) <- NA

CpGislands.gr <- suppressWarnings(resize(CpGislands.gr, 5000, fix="center"))


## calcute FDRs
stats <- calculateFDRs(m=meth.gr, CGIs=CpGislands.gr,PMDs = PMDsegments.gr,
                       num.cores=4)

FDR.cutoff <- 5 
m.sel <- 0.5 
n.sel=as.integer(names(stats$FDRs[as.character(m.sel), ]
                       [stats$FDRs[as.character(m.sel), ]<FDR.cutoff])[1])

## segment into UMRs and LMRs
UMRLMRsegments.gr <- segmentUMRsLMRs(m=meth.gr, meth.cutoff=m.sel,
                                     nCpG.cutoff=n.sel, PMDs = PMDsegments.gr,
                                     num.cores=10, myGenomeSeq=Hsapiens, 
                                     seqLengths=sLengths,pdfFilename = paste0(data_dir,"human_esc_hg19_chr2_segments.pdf"))

saveUMRLMRSegments(segs=UMRLMRsegments.gr,
                   GRangesFilename=paste0(data_dir,"UMRsLMRs.gr.rds"), 
                   TableFilename=paste0(data_dir,"UMRsLMRs.tab"))

```

### Find FMRs 

Since MethylSeekR does not produce FMR ranges we need to create them ourselves.

```{r,eval=FALSE}
## load methylseekR_hg19_chr2_umrlmr
seekr.esc <- readRDS(paste0(data_dir,"UMRsLMRs.gr.rds"))

seekr.esc
## no fmrs included so we take intermediate ranges as fmr
fmr <- gaps(seekr.esc)
fmr <- fmr[chrom(fmr)=="chr2"]
## remove stranded full chrom ranges
fmr <- fmr[-c(1,2)]

## we need addtional data as the mean methylation and #CG per segment 
mbw <- readRDS(file = meth.preload_chr2)
ovl <- findOverlaps(query = fmr, subject = mbw)
x_mean <- mclapply(unique(queryHits(ovl)), function(x) list(nCG = length(subjectHits(ovl[queryHits(ovl)==x])),pmeth=mean(mbw[subjectHits(ovl[queryHits(ovl)==x])]$score)),mc.cores = 10)
values(fmr) <- DataFrame(nCG = sapply(X = x_mean, function(x) x$nCG),pmeth=sapply(X = x_mean, function(x) x$pmeth),type="FMR")

## now we can "update" the orignal object 
seekr.esc.gr <- GRanges()
seekr.esc.gr <- c(seekr.esc.gr,fmr)
tmp <- seekr.esc
values(tmp)$nCG.segmentation <- NULL
values(tmp)$T <- NULL
values(tmp)$M <- NULL
values(tmp)$median.meth <- NULL
seekr.esc.gr <-  sort(c(seekr.esc.gr,tmp))
saveRDS(seekr.esc.gr,file = paste0(data_dir,"UMRsLMRsFMRs.gr.rds"))

```

## Segment using methylKit

Now we can use methylKit to segment the full methylome one more time.

```{r,eval=FALSE}

# library(methylKit)
set.seed(123)

mbw <- readRDS(meth.preload_full)

## we set the number of groups to be 3 
res = methSeg(mbw,diagnostic.plot=TRUE,maxInt=50,minSeg=5,G=3,eps=NULL)
saveRDS(res,file = paste0(data_dir,
                          "methylKit_hg19_methseg<maxInt=50,minSeg=5,G=3,eps=NULL>.rds"))
  
```

# Analyse data and create Figures

We can shortly have a look at the general methylation distributions.

```{r}
## load methylKit segments
res.esc <- readRDS(paste0(data_dir,
                          "methylKit_hg19_methseg<maxInt=50,minSeg=5,G=3,eps=NULL>.rds"))

res.esc <- res.esc[chrom(res.esc)=="chr2"]

hist(res.esc$seg.mean,xlab="methylation",col="firebrick",
     main="methylation values of segments",border="white")
plot(density(res.esc$seg.mean,bw=0.02))



## load MethylSeekR segments
seekr.esc.gr <- readRDS(paste0(data_dir,"UMRsLMRsFMRs.gr.rds"))

hist(seekr.esc.gr$pmeth,xlab="methylation",col="cornflowerblue",
      main="methylation values of segments",border="white")
plot(density(seekr.esc.gr$pmeth,bw=0.02))




```

And have a closer look on the distinct segment types.


```{r}

## split up GRanges by segment type
res.esc.grl <- split(res.esc,f = res.esc$seg.group)

seekr.esc.grl <- split(seekr.esc.gr,f = seekr.esc.gr$type)
# reorder to UMR < LMR < FMR
seekr.esc.grl <- seekr.esc.grl[c("UMR","LMR","FMR")]


# Compare the Methylation distribution per segment
res.esc.seglength <- lapply(res.esc.grl,FUN = function(x) log10(width(x)))
res.esc.meth <- lapply(res.esc.grl,FUN = function(x) x$seg.mean)
res.esc.cg <- lapply(res.esc.grl,FUN = function(x) log10(x$num.mark) )

seekr.esc.seglength <- lapply(seekr.esc.grl,FUN = function(x) log10(width(x)))
seekr.esc.meth <- lapply(seekr.esc.grl,FUN = function(x) x$pmeth)
seekr.esc.cg <- lapply(seekr.esc.grl,FUN = function(x) log10(x$nCG))


# require(reshape2)

feature.order <- c("1","UMR","2","LMR","3","FMR")
feature.names <- c("Segment Class 1","UMR","Segment Class 2","LMR","Segment Class 3","FMR")
feature.tool <- c("methylKit (3 Groups)","MethySeekR")
feature.colors <- c("#F8766D","#F8D56D")



all.seglength <- rbind(data.frame(melt(res.esc.seglength), Tool = "methylKit"),
                       data.frame(melt(seekr.esc.seglength),Tool = "MethylSeekR")
                       )
all.seglength$L1 <- factor(all.seglength$L1,levels = feature.order)

all.cg <- rbind(data.frame(melt(res.esc.cg), Tool = "methylKit"),
                  data.frame(melt(seekr.esc.cg),Tool = "MethylSeekR")
)
all.cg$L1 <- factor(all.cg$L1,levels = feature.order)

all.meth <- rbind(data.frame(melt(res.esc.meth), Tool = "methylKit"),
                       data.frame(melt(seekr.esc.meth),Tool = "MethylSeekR")
)
all.meth$L1 <- factor(all.meth$L1,levels = feature.order)

```

Now we overlap the segments with their functional annotation from Roadmap.

```{r}
## Annotate wih chromHMM

# H1 chromHMM 15 states 
chromhmm.file <- chromhmm_file
chromhmm.states <- read.delim(chromhmm_states,stringsAsFactors = FALSE)

chromhmm <- genomation::readGeneric(chromhmm.file,meta.cols = list(state=4))
chromhmm.grl <- split(chromhmm,f = chromhmm$state)
chromhmm.grl <- chromhmm.grl[gtools::mixedorder(names(chromhmm.grl))]

## annotate methylKit res.esc
seg.annot.res.esc <- annotateWithFeatures(target = res.esc.grl,features = chromhmm.grl)
seg.annot.res.esc.mat <- heatTargetAnnotation(seg.annot.res.esc,
                                              plot = FALSE,
                                              precedence = TRUE)

colnames(seg.annot.res.esc.mat) <- chromhmm.states$DESCRIPTION

# # now annotate methylseekr ranges
seg.annot.seekr.esc <- annotateWithFeatures(target = seekr.esc.grl,features = chromhmm.grl)
seg.annot.seekr.esc.mat <- heatTargetAnnotation(seg.annot.seekr.esc,
                                                plot = FALSE,
                                                precedence = TRUE)

colnames(seg.annot.seekr.esc.mat) <- chromhmm.states$DESCRIPTION


# require(reshape2)
mat <- rbind(seg.annot.res.esc.mat,seg.annot.seekr.esc.mat)
mat.m <- melt(mat,varnames = c("Feature","State"))
mat.m$Feature <- factor(mat.m$Feature,levels = c("1","UMR","2","LMR","3","FMR"))
```

And here we produce Figure 4 of the manuscript.

```{r}

# taken from http://stackoverflow.com/questions/12539348/ggplot-separate-legend-and-plot and 
# http://www.sthda.com/english/wiki/ggplot2-easy-way-to-mix-multiple-graphs-on-the-same-page-r-software-and-data-visualization
require(gridExtra)

get_legend<-function(myggplot){
  tmp <- ggplot_gtable(ggplot_build(myggplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)
}

# 1. Create the plots
#++++++++++++++++++++++++++++++++++
# Create box plot
require(ggplot2)
p_seglength <- ggplot(all.seglength, aes(y=value, x=L1,fill=Tool)) +
                  geom_boxplot(outlier.size = 0.1)+
                  coord_cartesian()+
                  labs(y="Segment length distributions\nlog10(bp)", x="Feature")+
                  scale_x_discrete(labels=feature.names)+
                  scale_fill_manual(values =feature.colors)+
                  theme(axis.text.x  = element_text(angle=45, vjust=1, hjust=1))
# p_seglength

p_cg <- ggplot(all.cg, aes(y=value, x=L1,fill=Tool)) +
            geom_boxplot(outlier.size = 0.1)+
            coord_cartesian()+
            labs(y="Number of CpGs per Segment\n log10 ", x="Feature")+
            scale_x_discrete(labels=feature.names)+          
            scale_fill_manual(values =feature.colors)+
            theme(axis.text.x  = element_text(angle=45, vjust=1, hjust=1),legend.position = "none")
#p_cg

p_meth <- ggplot(all.meth, aes(y=value, x=L1,fill=Tool)) +
          geom_boxplot(outlier.size = 0.1)+
          coord_cartesian()+
          labs(y="Methylation", x="Feature")+
          scale_x_discrete(labels=feature.names)+
          scale_fill_manual(values =feature.colors)+
          theme(axis.text.x  = element_text(angle=45, vjust=1, hjust=1),legend.position = "none")
# p_meth

p_heatmap <- ggplot(mat.m, aes(Feature, State)) + 
  geom_tile(aes(fill = value)) + 
  geom_text(aes(label = round(value, 1))) + 
  scale_fill_gradient(name="Percent\nOverlap",low = "white",high = "steelblue")+
  labs(y="Chromatin State", x="Feature")+
  scale_x_discrete(labels=feature.names)+
  theme(axis.text.x  = element_text(angle=45, vjust=1, hjust=1))
# p_heatmap


# 2. Save the legend
#+++++++++++++++++++++++
legend <- get_legend(p_seglength)
# 3. Remove the legend from the box plot
#+++++++++++++++++++++++
p_seglength <- p_seglength + theme(legend.position="none")

require(grid)
p_seglength <- arrangeGrob(p_seglength, top = textGrob("a", x=unit(0, "npc"),y=unit(1, "npc"),
                                                       just=c("left","top"), gp=gpar(col="black", fontsize=14)))

p_cg <- arrangeGrob(p_cg, top = textGrob("b", x=unit(0, "npc"),y=unit(1, "npc"),
                                                       just=c("left","top"), gp=gpar(col="black", fontsize=14)))

p_meth <- arrangeGrob(p_meth, top = textGrob("c", x=unit(0, "npc"),y=unit(1, "npc"),
                                                       just=c("left","top"), gp=gpar(col="black", fontsize=14)))

p_heatmap <- arrangeGrob(p_heatmap, top = textGrob("d", x=unit(0, "npc"),y=unit(1, "npc"),
                                                   just=c("left","top"), gp=gpar(col="black", fontsize=14)))



# 4. Arrange ggplot2 graphs with a specific width
#+++++++++++++++++++++++
pdf(file = paste0(figure_dir,"compTools.pdf"),width = 10,height = 7)
grid.arrange(p_seglength, p_cg, p_meth,legend,p_heatmap, nrow=2,ncol=4,layout_matrix = rbind(c(1,2,5,5),c(3,4,5,5)),widths=c(1, 1,1,1))
dev.off()

```

# SessionInfo

```{r}
sessionInfo()
```

